---
title: "Introduction to R for Finance"
output: html_notebook
---

### 3. Data Frames

#### Create and subset data frames
```{r financial returns and  present value of projected cash flows}
company <- c(rep("A", 3), rep("B", 4) )
cash_flow <- c(1000, 4000, 550, 1500, 1100, 750, 6000)
year <- c(1, 3, 4, 1, 2, 4, 5)

cash <- data.frame(company, cash_flow, year)  # create data frame
cash$present_value <- cash$cash_flow * 1.05^ -cash$year  # present value of each cash flow at a 5% interest rate
cash

# Total present value of cash
total_pv <- sum(cash$present_value)
paste("The total present value of cash is", total_pv)
cash_B <- subset(cash, cash$company == "B")  # subset company B
total_pv_B <- sum(cash_B$present_value)  # total present value of company B
paste("The total present value of cash of company B is", total_pv_B)
```



### 4. Factors

#### Create a factor

Bond credit ratings are common in the fixed income side of the finance world as a simple measure of how "risky" a certain bond might be. Here, riskiness can be defined as the probability of default, which means an inability to pay back your debts. 

This is a perfect example of a factor! It is a categorical variable that takes on a limited number of levels.

```{r}
# credit_rating character vector
credit_rating <- c("BB", "AAA", "AA", "CCC", "AA", "AAA", "B", "BB")
credit_factor <- factor(credit_rating)
str(credit_factor)
```



#### Factor levels
Accessing the unique levels of your factor is simple enough by using the **levels()** function. You can also use this to **rename** your factor levels!

```{r}
levels(credit_factor)  # identify unique levels
levels(credit_factor) <- c("2A", "3A", "1B", "2B", "3C")  # Rename the levels of credit_factor
summary(credit_factor)
plot(credit_factor)  # visualize factor
```

#### Bucketing a numeric variable into a factor...
using **cut()**

```{r}
AAA_rank <- sample(1:100, 50, replace = T)
AAA_factor <- cut(x = AAA_rank, breaks = c(0, 25, 50, 75, 100))  # use breaks to specify the groups
AAA_factor
levels(AAA_factor) <- c("low", "medium", "high", "very_high")  # rename the levels
AAA_factor
```

#### ordering and subsetting factors with ordered()

Create an ordered factor. By existing factor use **ordered()** to order it.
 
```{r}
credit_rating <- c("BB", "AAA", "AA", "CCC", "AA", "AAA", "B", "BB")
unique(credit_rating)  # Use unique() to find unique words
credit_factor_ordered <- factor(credit_rating, ordered = T, levels = c("AAA", "AA", "BB", "B", "CCC")) 
credit_factor_ordered 
credit_factor_ordered[c(-3, -5), drop = T]  # drop "A" and at the same time the level of "A" with drop = T
```

#### stringsAsFactors = F

R's default behavior when creating data frames is to convert all **characters** into factors.

To turn off this behavior, use stringsAsFactors = F

```{r}
credit_rating <- c("AAA", "A", "BB")
bond_owners <- c("Dan", "Tom", "Joe")
bonds <- data.frame(credit_rating, bond_owners)
str(bonds)
bonds <- data.frame(credit_rating, bond_owners, stringsAsFactors = F)
str(bonds)

# only create ordered factor for rating
bonds$credit_factor <- factor(bonds$credit_rating, ordered = T, levels = c("AAA", "A", "BB"))
str(bonds)
```


### 5. Lists

#### Create a list with list(

)
```{r}
name <- "Apple and IBM"
apple <- c(109.49, 109.90, 109.11, 109.95, 111.03)
ibm <- c(159.82, 160.02, 159.84, 160.35, 164.79)
cor_matrix <- cor(cbind(apple, ibm))
portfolio <- list(name, apple, ibm, cor_matrix)
names(portfolio) <- c("portfolio_name", "apple", "ibm", "correlation")  # add names to the list
portfolio
```

#### Access elements in a list

To access the elements in the list, use [ ]. This will always return another list.

```{r}
portfolio[2]
```

To pull out the data inside each element of your list, use [[ ]].
```{r}
portfolio[[2]]
```

If your list is named, you can use the $ operator. This is the same as using [[ ]] to return the inner data.

```{r}
portfolio$apple
```

### Adding to and removing from a list

You can do so using **$** like you did when adding new columns to data frames.

```{r}
# add
portfolio$weight <- c(apple = 0.2, ibm = 0.8)
portfolio

# edit value
portfolio$weight <- c(apple = 0.3, ibm = 0.7)
portfolio

# remove by setting to NULL
portfolio$microsoft <- c(150.0, 152, 154, 154.5)
portfolio
portfolio$microsoft <- NULL
portfolio
```


### split() and unsplit()

Often, you will have data for multiple groups together in one data frame. We split up data frame into separate data frames divided by a variable.

```{r}
cash
# split cash into a list of 2 data frames separated by company
split_cash <- split(cash, f = cash$company)  
split_cash

# To get your original data frame back, use unsplit()
original_cash <- unsplit(split_cash, f = cash$company)
original_cash
```


#### Split-Apply-Combine
A common data science problem is to split your data frame by a grouping, apply some transformation to each group, and then recombine those pieces back into one data frame.

```{r}
split_cash <- split(cash, f = cash$company)

# Print the cash_flow column of B in split_cash
split_cash$A$cash_flow 
split_cash$A$cash_flow <- 0

# Use the grouping to unsplit split_cash
cash_edit <- unsplit(split_cash, f = cash$company)
cash_edit

```

#### Attributes

Attributes are a bit of extra metadata about your data structure. Some of the most common attributes are: **row names and column names, dimensions, and class**. You can use the **attributes()** function to return a list of attributes about the object you pass in. To access a specific attribute, you can use the **attr()** function.

```{r}
attributes(cash)
attr(cash, "names")  # access a specific attribute
```

